plyranges: a grammar for transforming genomics data

### 2018-04-18 paper.Rmd
algebra stuff from wiki - will rework

>>>>

## Genomic Relational Algebra

* Arithmetic
** Unary
   IRanges has already explored this space and defined its own
   "algebra", but we could revisit it. bedtools is also inspiring.

   We need to make adjustments to the start, end and width. Extending
   dplyr, this would just be =mutate()=, e.g.:
   #+begin_src R
   mutate(x, start=start+1L)
   #+end_src
   And while that could be supported, we typically want to make
   relative adjustments.

   One complication is that the start, end and width are mutually
   dependent. Changing the start, as above, needs to change either
   the width or the end to preserve integrity. We assume that the
   user thinks of the range as a pair of endpoints (this is natural
   in genomics anyway), so setting the start will modify the width
   but preserve the end. The opposite, preserving the start but not
   the other endpoint, is =IRanges::shift()=. Maybe we should be more
   explicit about whether the width or the other endpoint is changed
   by side effect.

   Here are some possible verbs for changing the width:
   * =stretch_start/end= :: Extend the range by 'x' by pulling the
	start or the end (and leaving the other one anchored),
   * =squeeze_start/end= :: Opposite of stretch (could just use
	negative value for stretch, but being explicit seems
	preferable in many cases),

   Not sure whether =squeeze= is actually useful, as we typically
   want to stretch a range to consider context, not squeeze it.

   Or maybe it would be more obvious if we supported the notion of
   anchoring, similar to grouping. This would require decorating the
   object. For example,
   #+begin_src R
     stretch(anchor_end(x), y)
   #+end_src
   This pushes information from the syntax to the object state. While
   it simplifies the grammar, object behavior is more complex and
   implicit. But if the code is part of the same chain, it seems
   intuitive.

   There would be three ways to anchor:
   * =anchor_start()=
   * =anchor_end()=
   * =anchor_center()=

   Anchoring the midpoint would enable symmetric zooming.

   This is basically a deconstruction of =IRanges::resize()=, except
   =resize()= is absolute. Perhaps =set_width()= could play
   the role of =resize()=. It is not clear whether we need
   =set_start()= and =set_end()=.

   When strand is involved, we will need something like:
   * =anchor_3p= (start for negative strand)
   * =anchor_5p= (end for negative strand)

   Anchoring the width would not make sense, because (1) it is not a
   point that can be intuitively anchored and (2) stretching and
   squeezing obviously modify the width, so it would not apply to
   those.

   For width-invariant changes, there is =shift()= from the IRanges
   API. That is a symmetric operation (positive for right, negative
   for left). We could introduce:
   * =shift_left()=
   * =shift_right()=
   If we wanted the direction to be explicit.

   Note that with GRanges, strand can modulate direction, and we need
   to distinguish left/right (chromosomal coordinates) from the
   direction of transcription:
   * =shift_upstream()=
   * =shift_downstream()=

     | Fluent                   | IRanges       | bedtools |
     |--------------------------+---------------+----------|
     | stretch(anchor_start())  | -             | slop -r  |
     | stretch(anchor_end())    | -             | slop -l  |
     | stretch(anchor_center()) | +()           | slop -b  |
     | set_width(anchor_*)      | resize(fix=*) | -        |
     | shift_right()            | shift()       | shift -s |
     | shift_left()             | -             | -        |
     | shift_up/downstream()    | -             | -        |

  Note that =resize,GRanges()= and =slop -s= use the strand for
  direction, i.e, =anchor_3p()= and =anchor_5p()=. Bedtools supports
  width-relative adjustment with =slop -pct=; we will let the user
  be explicit, e.g.: =stretch(x, 2*width(x))=.

  One problem is that can be stretched or shifted out of bounds, or squeezed out of existence.
  Currently, GenomicRanges will warn the user when a range is pushed
  out of bounds and suggest calling =trim()=. We should probably
  throw an error in this case, because it is never a good idea for
  an object to be in a (semi)-invalid state. The resolution needs
  to be explicit, as part of the function call. By default,
  we throw an error. There could be variants of all of these
  operations with the post fix =_up_to= that imply the trim, 
  e.g., =stretch_up_to()= or =set_width_up_to()=. Another
  resolution would be to drop the ranges that would become
  out of bounds. Those could use the =_or_drop()= postfix.

  We could also use this anchoring notion to define a new verb
  =flip()= that generates flanking regions, but that will cause
  semantic issues with strandedness. Instead, we probably need
  =flank_left()= and =flank_right()=. With strand, there would be
  =flank_before()= and =flank_after()=.

** Binary
   Binary operations between two ranges mostly align with set
   operations that treat ranges as sets of integers. These are
   parallel/vectorized operations between two vectors/tables of
   ranges, so are not to be confused with the aggregating set
   operations described later.

   Currently, these set operations have the =p= prefix, e.g.,
   =punion()=. We could make the binary nature more obvious by
   aliasing them as operators:
   * =%union%=
   * =%intersect%=
   * =%setdiff%=

   There are also operations that relate to the relative positions of
   two ranges:
   * =between()=: the range in the gap between the two ranges
     * To get the separation, just do =width(between(x, y))=
   * =span()=: a range that spans both ranges (and the gap)

   As a note, to express a binary operation involving the subject of a
   =mutate()=, =summarize()= or whatever call, we could use the pronoun:
   #+begin_src R
   mutate(x, foo = width(. %intersect% other_ranges))
   #+end_src
